Measuring Individual Partisan Harm in Redistricting

### 2021-10-12 paper/paper.Rmd
round 1 edits

We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R,
    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
$$for $g:[0,1]\bijto\R$ an increasing function.

@@

Then we may define the utilities for a (randomly selected) voter as $$
    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
    U^{(L)} \dfeq u^{(L)}(I, V, Q).
$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.

@@

Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis]. **[CM: revisit ¶ based on NJ findings]****.**

```{=html}
<!--  
With Harm defined, we can thus define Partisan Equity as: $E[H|V(I) = DEM] = E[H|V(I) = REP]$

Thus Fairness can be described as the signed movement away from equity. This naming works because it's condition on the votes and geographic distribution that you have, rather than some neutral midpoint that would be hard to justify.

This clarifies our QOI and by later focusing on fairness, that gives us the analogous connection between bias (symmetry) and fairness (equity).

Later, that also gives room in the discussion of the EGAP to talk about why you could describe a utility function based on $Pr(Vote not wasted)$, which 
-->
```
As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.

@@

## Differential Harm and Fairness

Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
This is easily expressed using our notation above; $Q$ is fair in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.

To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
  \Delta^{p_1}_{p_2}H(q) \dfeq = \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
$$ for two parties $p_1$ and $p_2$.
So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta H^\dem_\rep(q)$.
But the same concepts and definitions naturally carry over to the multiparty setting.

@@

The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H^(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.

@@

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace with rate 2.[^3]
Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

@@

A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].

>>>>

We posit that the utility of the global election outcome to a voter is a strictly increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R,
    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
$$for $g:[0,1]\bijto\R$ a strictly increasing function.

@@

Then we may define the utilities for a (randomly selected) voter as $$
    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
    U^{(L)} \dfeq u^{(L)}(I, V, Q).
$$Thus, the local utility for a single voter $i$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.

@@

Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
As we discuss below in Section \ref{subsec:diff-harm}, locally-based harm is connected to global electoral outcomes, anyway.

As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.

@@

## Differential Harm and Fairness {#subsec:diff-harm}

Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
This is easily expressed using our notation above; $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.

To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
  \Delta^{p_1}_{p_2}H(q) \dfeq \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
$$ for two parties $p_1$ and $p_2$.
So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
But the same concepts and definitions naturally carry over to the multiparty setting.

@@

The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.

@@

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

@@

A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
These concerns to not apply to our proposed measures.
